Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-23 latex/main.tex
Updates from Overleaf

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with 
% Not sure about this assertion! Yan?
samples separated by millions of years of evolution.
[This is rough outline text]
Complex models are required to fill in the gaps in our knowledge,
incorporating prior information about sequence evoltion etc etc.
In this context, recombination can be treated as a nuisance factor 
because we cannot hope to observe it over these timescales [or something
similar].
Phylogenetics has continued along this vein, 
although with various challenges on how to accomodate 
more closely related samples and to handle potential recombination met.

The data collected during the COVID-19 pandemic as part of global
surveillance fundamentally changed these assumptions. 
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. This superabundance of data 
overwhelmed classical phylogenetic methods, both in terms 
of the number of sequences that could be processed, but also
more fundamentally in terms of the underlying assumption of 
retrospective data sparsity. [TODO rephrase]
The UShER method \cite{Turakhia2021} conclusively demonstrated 
the value of simpler models that build a tree iteratively
in real-time, quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.

Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Despite this, phylogenies used as the basis for innumerable studies 
continue to depict the genealogy of SARS-CoV-2 as a single
tree~\cite{McBroome2021,Hadfield2018}.

% The genomic revolution has transformed how we study the evolution and spread of pathogens,
% allowing real-time tracking of pathogens over time and space.
% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
% quickly overwhelming typical phylogenetic approaches.
% Although real-time phylogenetic methods now exist to cope with this flood of genomic data

% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
% we lack methods to build phylogenies that can integrate recombination
% in the face of ongoing sampling of infected cases.

% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
% UShER has generated phylogenies updated daily \cite{McBroome2021} and
% has proven to be a powerful resource that successfully enabled the world
% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
% These phylogenies are used to identify and monitor
% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
% which might have acquired features to become Variants of Concern (VOC),
% for example, increased transmissibility, disease severity, and immune escape.
% Beyond supporting real-time genomic epidemiology,
% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
% UShER uses a tree to represent evolutionary change via mutation,
% assuming no recombination \cite{Turakhia2021}.

@@

@@

\subsection{Sc2ts recombination detection results for additional Pango X lineages}

\begin{table}
\centering
\caption{
Sc2ts recombination detection results for Pango X lineages not added to the ARG.
}
\label{tab:absent_pango_x}
\begin{tabular}{crrr}
\toprule
Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
\midrule
XBJ & 2 & 1 & 1\\
XBP & 8 & 2 & 1\\
XBS & 19 & 15 & 1\\
XBW & 1 & 1 & 1\\
XCA & 11 & 4 & 1\\
XAK & 2 & 2 & 1\\
XAY & 15 & 11 & 6\\
XBC & 36 & 21 & 3\\
XBL & 2 & 2 & 0\\
XBT & 1 & 1 & 2\\
\bottomrule
\end{tabular}
\end{table}

>>>>

\section*{INTRODUCTION}

%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
%%% and nonstandard abbreviations should be defined when first used in the text.

The genomic revolution has transformed how we study the evolution and spread of pathogens,
allowing real-time tracking of pathogens over time and space.
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
quickly overwhelming typical phylogenetic approaches.
Although real-time phylogenetic methods now exist to cope with this flood of genomic data
(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
we lack methods to build phylogenies that can integrate recombination
in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
UShER has generated phylogenies updated daily \cite{McBroome2021} and
has proven to be a powerful resource that successfully enabled the world
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
These phylogenies are used to identify and monitor
potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
which might have acquired features to become Variants of Concern (VOC),
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology,
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
UShER uses a tree to represent evolutionary change via mutation,
assuming no recombination \cite{Turakhia2021}.
However, given extensive evidence for recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
the tree-like representation of UShER is no longer tenable.

@@

% TODO merge this in
Ancestral Recombination Graphs (ARGs) are reticulate genealogies
that incorporate both mutation and recombination events \cite{Wong2024},
potentially allowing more comprehensive evolutionary representations of SARS-CoV-2
and other viruses undergoing recombination.
Until recently, ARGs lacked software support and
sufficiently scalable inference methods to be of practical use.
Recent advances enable ARG inference for tens to hundreds of thousands of human genomes \cite{Kelleher2019,Speidel2019,Zhang2023},
with high levels of recombination.
These advances now enable evolutionary genealogies to be built quickly and at scale, and
there is a surge of excitement in the inference and applications of ARGs \cite{Brandt2024,Lewanski2024,Wong2024,Nielsen2025}.

@@
